Supporting Information

### 2021-09-02 vertical-thermal-review.Rmd
Table 1: fixed up formatting and knit into doc
@NidhiVinod , I tweaked the formatting in your word doc, printed as pdf, created .pngs, and knit those into the doc. I also copied in the references list. After all that nocite confusion, it turns out it will be easier to just put the references in the list in the .Rmd!

Co-Authored-By: Nidhi Vinod <62031681+NidhiVinod@users.noreply.github.com>

![Table `r table_leaf_traits`. Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`. (*For latest version of table, see tables.pdf.*)](table1.png)

@@

## Implications [ & future research directions?]

@@

### warming

Warming impacts forests, from leaf metabolism to whole forest structure. Leaf level $T_{opt}$ is closely related ecosystem $T_{opt}$. Temperature limitations to leaf processes, such as VPD and gs-limitation, can modify ecosystem $T_{opt}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima. Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021]. A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018], especially if RH in the understory does not inhibit transpirational cooling.  Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019]. Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [McDowell et al. 2018; REFS] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].

### canopy disturbance

Canopies are crucial to microclimatic stratification, as they buffer several aspects of forest microenvironments [Figs 1-2, @zellweger_seasonal_2019]. This buffering is influenced by hydrology, as wetter sites and seasons allow for denser, and more buffered canopies  [@davisMicroclimaticBufferingForests2019]. Severe droughts and rising temperatures place canopies at risk, as larger trees tend to suffer more during droughts [@bennettLargerTreesSuffer2015]. The disproportionate impact of disturbance on large trees holds for other types of disturbance, including forest fragmentation and logging [[@lauranceRainForestFragmentation2006; @millerReducedImpactLogging2011], although fire tends to predominantly affect smaller trees [@brandoDroughtsWildfiresForest2019]. Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]

>>>>

\newpage
**Table `r table_leaf_traits`.** Studies listed here were compiled using a systematic review process, as described in Supplementary Information Methods `r appendix_lit_review_methods`.

![](table1a.png)

\newpage
Table `r table_leaf_traits`, cont.

![](table1b.png)

*1. Mau et al. 2018; 2. Coble and Cavaleri 2014; 3. Sack et al. 2006; 4. Chin and Sillett 2019; 5. Wyka et al. 2012; 6. Atherton et al. 2017; 7. Kenzo et al. 2015; 8. Kusi and Karasi 2020; 9. Dang et al. 1997; 10. Gebauer et al. 2015; 11. Marenco et al. 2017; 12.Kafuti et al. 2020; 13. Van Wittenberghe et al. 2012; 14. Zhang et al. 2019; 15. Weerasinghe et al. 2014; 16. Oldham et al. 2010; 17. Ichie et al. 2016; 18. Gregoriou et al. 2007; 19. Levizou et al. 2005; 20. Liakoura 1997; 21. Fauset et al. 2018; 22. Niinemets et al. 1998, 23. Ishida et al. 1998; 24. Millen and Clendon 1979; 25. Smith and Carter, 1988; 26. Hadley and Smith 1987; 27. Panditharathna et al. 2008; 28. Baltzer and Thomas 2005; 29. Dietz et al. 2007; 30. Coble et al. 2016; 31. Scartazza et al. 2016; 32. Duursma and Marshall, 2006; 33. Harley et al. 1996;  34. Hernandez et al. 2020; 35. Turnbull et al. 2003; 36. Chen et al. 2020; 37. van de Weg et al. 2012; 38. M.A Cavaleri et al. 2008;  39. Koniger et al. 1995; 40. Mastubara et al. 2009; 41. Harris and Medina 2013; 42. Hansen et al. 2001; 43. Poorter et al. 1995; 44. Coble et al. 2017; 45. Niinemets et al. 2004; 46. Poorter et al. 2000
```{r eval=FALSE, echo=FALSE}
# refs in this table need to go in the nocite field at top of document 
```

@@

## Implications

@@

### Warming

Warming impacts forests, from leaf metabolism to whole forest structure. Leaf level $T_{opt}$ is closely related ecosystem $T_{opt}$. Temperature limitations to leaf processes, such as VPD and gs-limitation, can modify ecosystem $T_{opt}$ if species do not acclimate, resulting in reduced carbon sequestration [@huangAirTemperatureOptima2019; @wayJustRightTemperature2019; @tanOptimumAirTemperature2017; @bennettThermalOptimaGross2021]. At a metabolism level, leaf photosynthesis and respiration are coupled to photosynthetic thermal optima. Beyond this inflection point, photosynthesis declines whereas respiration continues to increase exponentially, eventually shifting the carbon balance from sink to source, independent of water and irradiance [@duffyHowCloseAre2021]. Exposed canopies are likely most vulnerable, as current growing season temperatures are equal to, or higher than photosynthetic optima of temperature and tropical forests [@slotSituTemperatureResponse2017; @mauTemperateTropicalForest2018; @huangAirTemperatureOptima2019; @kumarathungeAcclimationAdaptationComponents2019], and leaf temperatures may  approach lethal temperatures [e.g., @osullivanThermalLimitsLeaf2017; @tiwariPhotosyntheticQuantumEfficiency2021]. A buffering effect from the canopies might allow shaded layers to photosynthesize longer in the day [@millerOnlySunlitLeaves2021; @heChangesShadowShifting2018], especially if RH in the understory does not inhibit transpirational cooling.  Thermal limits are especially narrow in tropical and mid-latitudes forests [@osullivanThermalLimitsLeaf2017]. A daytime temperature threshold of 32 degrees C has been suggested for long term climate sensitivity of global tropical forests, where reduced growth, and increased drought and tree mortality will lead to increased carbon source than sink [@sullivanLongtermThermalSensitivity2020].
In the understory, warming is disproportionately affecting the less thermally-adapted plant species, resulting in associated plant community thermophilization [@zellwegerForestMicroclimateDynamics2020; @duqueThermophilizationAdultJuvenile2015; @greiserHidingClimateCharacterizing2019]. Canopy disturbance (see next section) along with a trend towards younger forests and again reduced tree growth associated with warming [@mcdowellPervasiveShiftsForest2020] poses an increasing threat to the biodiversity of microrefugia that are otherwise buffered from extremes [@scheffersMicrohabitatsReduceAnimal2013; @greiserHidingClimateCharacterizing2019].

### Canopy disturbance

Canopies are crucial to microclimatic stratification, as they buffer several aspects of forest microenvironments [Figs 1-2, @zellweger_seasonal_2019], particularly in wetter environments with dense canopies [@davisMicroclimaticBufferingForests2019]. Severe droughts and rising temperatures place canopies at risk, as larger trees tend to suffer more during droughts [@bennettLargerTreesSuffer2015]. The disproportionate impact of disturbance on large trees holds for other types of disturbance, including forest fragmentation and logging [[@lauranceRainForestFragmentation2006; @millerReducedImpactLogging2011], although fire tends to predominantly affect smaller trees [@brandoDroughtsWildfiresForest2019]. Canopy structure affects understory species composition, which has been shown to shift under warming and canopy disturbance [@majasalmiImpactTreeCanopy2020; @bertrandCommentForestMicroclimate2020; @maesPlantFunctionalTrait2020]
